Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-02 method.tex
Notation update work in progress
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@846 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
search with $\sim$minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$%
%
\editorial{Change this to a harpoon or vector symbol.}
%
that determine the amplitude and phase evolution of the gravitational wave strain. For systems where the effects of spin can be ignored, the
intrinsic parameters are the component masses of the binary, $\bar{\theta} =
(m_1, m_2)$. Searches for gravitational waves from compact binary coalescence
typically employ matched filter banks that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  The filters
for the waveforms $h(t,\bar{\theta})$, known as templates, are the waveforms
weighted by the inverse detector noise amplitude spectral density. To construct a template bank, templates are chosen with discrete signal parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}. That is, any possible
signal within the search space will have a cross-correlation of $\geq0.97$ with
at least one template. Such a template bank is said to have a {\em
minimum match} of 0.97. Data are filtered against each template to produce an \SNR{}
time-series. Local peak-finding across time and templates determines detection
candidates.
%
\footnote{There are two gravitational-wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work, as it is
straightforward to generalize but not necessary for understanding any of the
points that will be made.}

We will denote the $i^\mathrm{th}$ filter with parameters $\bar{\theta_i}$
as a function of time $x_i(\tau)$.
In this work we will discuss transformations to a set of filters
$\{x_i(\tau)\}$.  Some of these transformations are not useful or practical
over the entire parameter space.  For that reason we assume from here onward
that the set of filters $\{x_i(\tau)\}$ refer to a set of near-neighbor filters
that can be chosen as a subset of the full parameter space.  Several such
subsets can be chosen until all of the filters are a member of one local set.
\editorial{nvf: I don't think that the splitting of the bank is a crucial point here. Perhaps later if we discuss the up-front cost of the SVD.}

Filtering the detector data $h(t)$ involves a convolution of the data with the
filter.  For a unit-normalized filter, and whitened detector data, the result
can be interpreted as the signal-to-noise ratio, $\rho_i(t)$ and is defined as
%
% Filtering equations
%
\begin{eqnarray}
\rho_i(t) &=& \int x_i(\tau) \, h(t-\tau) \, d \tau \label{eq:SNRTD} \\
          &=& \int \tilde{x}_i(f) \, \tilde{h}(f) \, \e^{-2\pi\i ft} \label{eq:SNRFD} df,
\end{eqnarray}
%
%
where the second line is a result of the convolution theorem and
$\tilde{x}_i(f)$ is the Fourier transform of $x_i(\tau)$ as is $\tilde{h}(f)$ the
Fourier transform of $h(t)$.

The evaluation of the integrals in \eqref{eq:SNRTD} and \eqref{eq:SNRFD} are
implemented as sums over sample points for the digitized gravitational-wave
detector output.  Discrete Fourier transforms can be computed efficiently
numerically.  For that reason \eqref{eq:SNRFD} is typically far faster,
computationally.  To evaluate \eqref{eq:SNRTD} requires \order{\tmpsamps
\hoftsamps} floating point operations per filter $x_i$, where \tmpsamps\ is the
number of sample points in the filter $x_i$ and \hoftsamps\ is the number of
sample points in the data $h$.  Assuming \numtmps\ filters are required, the
total cost is \order{\numtmps \tmpsamps \hoftsamps} However, \eqref{eq:SNRFD}
requires only \order{\hoftsamps \log \hoftsamps} operations per filter assuming
transform lengths that are longer than the filter (i.e. \hoftsamps\ $>$
\tmpsamps), resulting in a total cost of \order{\numtmps \hoftsamps \log
\hoftsamps}.  In most cases \tmpsamps\ $\gg$ $\log$ \hoftsamps\ and the
computational savings by choosing the frequency-domain integral form
\eqref{eq:SNRFD} is clear.  However, to take full advantage of the
computational efficiency of \eqref{eq:SNRFD} requires an acausal knowledge of
the detector data $h(t)$, which implies an inherent latency.  In contrast,
\eqref{eq:SNRTD} can be updated every time a new sample point of detector data
is taken.

@@

In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
\rho_i(t) \approx \sum_k^{\numslices} \sum_j^{\numsvdtmps} 
	\int_{\tau_{k}}^{\tau_{k+1}} v_{ijk} \sigma_{jk} u_{jk}\,(\tau) \, h(t-\tau) \, d \tau \label{eq:decomp}
\end{equation} 
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

The first step of the orthogonal decomposition described in \eqref{eq:decomp}
is to divide the templates into \emph{time slices}.  This is a time-domain
analogue to the frequency-domain decomposition described in ~\cite{Marion2004,
Buskulic2010, beauville2006, beauville2008}.  A matched filter is constructed
for each time slice.  The outputs form an ensemble of partial \SNR{}
streams.  By linearity, these partial \SNR\ streams can be suitably time
delayed and summed to reproduce the \SNR\ of the full template.  We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t_0$, or equivalently frequency \fmax.
This is often chosen to correspond to the \ISCO\ defined previously. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t_\mathrm{0})$ and is said to have  a
duration of $t_\mathrm{0} - t_\mathrm{low}$. It is critically sampled at a
rate of $2 \fmax$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time-slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

Rather than applying a unique time-slice decomposition to each template
waveform, we find a decomposition that is adequate for the entire set
$\{x_i\}$.  This is facilated by choosing templates in the set $\{x_i\}$ with
similar chirp masses, $\mathcal{M}$.  The time-slice decompositions of the
filters $x_i(\tau)$ lead to new filters $y_{ik}(\tau)$ satisfying
%
\begin{equation}
\label{eq:time_slices}
x_i(\tau) \approx \sum_k^{\numslices} y_{ik}(\tau)
\end{equation}
where \numslices\ is the number of time slices required.  Note that we write
the relationship as approximate since the resampling implementation may have
some inherent loss in quality.  We note that by construction these filters are
orthogonal over the index $k$ since they are disjoint in time.  In the next
section we examine how to reduce the number of filters $y_{ik}(\tau)$ via
singular value decomposition to construct a set of filters that is also
orthogonal within a given time slice.

@@

We now examine the computational cost scaling of the approximate implementation
of \eqref{eq:SNRTD} as \eqref{eq:decomp}.  An actual implementation of this
decomposition in a working analysis pipeline is discussed in the next section
along with measured computational requirements.  For convenience, table
\ref{tab:recap} is a recap of the meaning of various symbols used in this
calculation.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\begin{tabular}{rl}
\bf{Symbol}	& \bf{Definition} \\
\hline
\hoftsamps	& The number of sample points in the data \\
\tmpsamps	& The number of sample points in template $x_i$ \\
\numtmps	& The number of templates in the set $\{x_i\}$ \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal filters in time slice $k$ \\
\slicessamps	& The number of samples in time slice $k$ \\
\fmax		& The maximum frequency of a filter \\
$f_k$		& The sample frequency of the $k^{\mathrm{th}}$ filter \\
\resampsamps	& The number of sample points in the resample filter
\end{tabular}
\caption{\label{tab:recap} Notation used to describe filtering.  This table
provides a quick reference for symbols used.}
\end{table}

@@

The filter bank can be implemented using finite impulse response (\fir{}) filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

@@

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time-slice layout \dots}
\begin{table}[!h]
\label{table:comp_cost}
\begin{center}
\input{flop_budget_example}
\end{center}
\caption{\hl{FIXME: Operation counts per sample for four different detection methods.}}
\end{table}

>>>>

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in a
search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source parameters
$\vec\theta$ that determine the amplitude and phase evolution of the gravitational wave strain. For systems where the effects of spin can be ignored, the
intrinsic source parameters are the component masses of the binary, $\vec\theta =
(m_1, m_2)$. Searches for inspiral signals
typically employ matched filter banks that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  For a given source, the strain observed by the detector is a linear combination of two waveforms corresponding to the `$+$' and `$\times$' gravitational wave polarizations, so for any value of $\vec\theta$ we must implement 2 filters.  The coefficients for the $M$ filters
are known as templates, and are formed by discretizing and time reversing the waveforms and weighting them by the inverse amplitude spectral density of the detector's noise. To construct a template bank, templates are chosen with discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\, \vec\theta_{M/2-1}$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}. That is, any possible
signal within the a given range of intrinsic source parameters will have an inner product that is $\geqslant 0.97$ with
at least one template. Such a template bank is said to have a {\em
minimum match} of 0.97. The data from the detector are whitened and convolved with each template to produce $2M$ \SNR\
time series. Local peak-finding across time and templates determines detection
candidates.

Filtering the detector data involves a convolution of the data with the
template.  For a unit-normalized template $h_i[k]$ and whitened detector data $x[k]$, both sampled at a rate $f^0$, the result can be interpreted as the signal-to-noise ratio, $\rho_i[k]$ defined as
%
% Filtering equations
%
\begin{equation}
	\label{eq:SNRTD}
	\rho_i [k] = \sum_{n=0}^{N-1} h_{i}[n] x [k-n].
\end{equation}

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain as an \textsc{fir} filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point operations per sample.  However, it is typically much more computationally efficient to use the convolution theorem and the \textsc{fft} to implement fast convolution, requiring only $\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample.

@@

\editorial{This paragraph is hereby condemned!  It's too early to introduce this equation, and this should be more of a summary of our method.}In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
\rho_i(t) \approx \sum_k^{\numslices} \sum_j^{\numsvdtmps} 
	\int_{\tau_{k}}^{\tau_{k+1}} v_{ijk} \sigma_{jk} u_{jk}\,(\tau) \, h(t-\tau) \, d \tau \label{eq:decomp}
\end{equation} 
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

Our first innovation and the first step of our proposed method
is to divide the templates into \emph{time slices} in a time-domain
analogue to the frequency-domain decomposition described in ~\cite{Marion2004,
Buskulic2010, beauville2006, beauville2008}\editorial{Four citations is way too many!  We should pick just \emph{one} citation for \textsc{mbta}, and it should be the most recent and complete description of it.}.  A matched filter is constructed
for each time slice.  The outputs form an ensemble of partial \SNR{}
streams.  By linearity, these partial \SNR\ streams can be suitably time
delayed and summed to reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\textrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$s^{\mathrm{th}}$ filter.}
\end{table}
%

\editorial{Need to convert some continuous time notation to discrete time here.}Rather than applying a unique time-slice decomposition to each template
waveform, we find a decomposition that is adequate for the entire set
$\{x_i\}$.  This is facilated by choosing templates in the set $\{x_i\}$ with
similar chirp masses, $\mathcal{M}$.  The time-slice decompositions of the
filters $x_i(\tau)$ lead to new filters $y_{ik}(\tau)$ satisfying
%
\begin{equation}
\label{eq:time_slices}
x_i(\tau) \approx \sum_k^{\numslices} y_{ik}(\tau)
\end{equation}
where \numslices\ is the number of time slices required.  We note that
the time slice decomposition in equation~(\ref{eq:time-slices}) is manifestly
orthogonal since the time slices are disjoint in time.  In the next
section we examine how to reduce the number of filters within each time slice via
singular value decomposition of the time sliced templates.

@@

We now examine the computational cost scaling of the approximate implementation
of \eqref{eq:SNRTD} as \eqref{eq:decomp}.  An actual implementation of this
decomposition in a working analysis pipeline is discussed in the next section
along with measured computational requirements.  For convenience, table
\ref{tab:recap} is a recap of the meaning of various symbols used in this
calculation.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\begin{tabular}{rl}
\bf{Symbol}	& \bf{Definition} \\
\hline
\tmpsamps	& The number of sample points in template $x_i$ \\
\numtmps	& The number of templates in the set $\{x_i\}$ \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal filters in time slice $k$ \\
\slicessamps	& The number of samples in time slice $k$ \\
\fmax		& The maximum frequency of a filter \\
$f^s$		& The sample frequency of the $k^{\mathrm{th}}$ filter \\
\resampsamps	& The number of sample points in the resample filter
\end{tabular}
\caption{\label{tab:recap} Notation used to describe filtering.  This table
provides a quick reference for symbols used.}
\end{table}

@@

\begin{comment}
The filter bank can be implemented using finite impulse response (\fir{}) filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

@@

\end{comment}
